---
title: "danl"
editor: visual
jupyter: python3
---

```         
```

```{r}

# Load libraries
library(tidyverse)
library(sf)
library(socviz)
library(scales)

# Load election data
election_panel <- read_csv(
  "https://bcdanl.github.io/data/election_panel_2000_2024.csv"
)

# Load county geometries
counties <- socviz::county_map |>
  st_as_sf()

# Join election data to map
map_data <- counties |>
  left_join(election_panel, by = c("id" = "id"))

# Plot
ggplot(map_data) +
  geom_sf(aes(fill = pct_DEMOCRAT), color = NA) +
  
  facet_wrap(~year, ncol = 3) +
  
  scale_fill_gradientn(
    colors = c("#FFFFFF", "grey80", "#81ABD9", "#211f82"),
    values = scales::rescale(c(0, 0.25, 0.5, 1)),
    name = "Percent Democrat",
    labels = percent_format(scale = 1)
  ) +
  
  labs(
    title = "U.S. Presidential Elections, 2000–2024",
    subtitle = "Votes toward Democratic Presidential Candidates"
  ) +
  
  theme_void() +
  
  theme(
    strip.background = element_rect(fill = "#6C8EBF", color = NA),
    strip.text = element_text(color = "black", face = "bold"),
    
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 12),
    
    legend.position = "bottom"
  )



```

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(scales)
library(ggthemes)
library(hrbrthemes)
library(rmarkdown)
library(socviz)

# Load data
election_panel <- read_csv(
  "https://bcdanl.github.io/data/election_panel_2000_2024.csv"
)

# County map data
county_data <- socviz::county_map

# Join data
map_data <- county_data %>%
  left_join(election_panel, by = c("fips" = "county_fips"))

# Plot
ggplot(map_data, aes(long, lat, group = group, fill = dem_share)) +
  geom_polygon(color = NA) +
  
  # Facet by year
  facet_wrap(~year, ncol = 3) +
  
  # Your exact colors (continuous gradient)
  scale_fill_gradientn(
    colors = c("#FFFFFF", "#211f82", "grey80", "#81ABD9"),
    na.value = "grey80"
  ) +
  
  coord_quickmap() +
  
  labs(
    title = "U.S. Presidential Elections, 2000–2024",
    subtitle = "Votes toward Democratic Presidential Candidates",
    fill = "Percent Democrat"
  ) +
  
  theme_void() +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12)
  )
```

```{r}
library(tidyverse)
library(socviz)

# Load data
election_panel <- read_csv(
  "https://bcdanl.github.io/data/election_panel_2000_2024.csv"
)

# Pick a year (example: 2020)
df <- election_panel %>%
  filter(year == 2020)

# Join to county map
county_df <- county_map %>%
  left_join(df, by = c("id" = "fips"))

# Your required colors:
# White → "#FFFFFF"
# Dark blue → "#211f82"
# Grey80 → "grey80"
# Light blue → "#81ABD9"

# Build a 4‑color continuous gradient
my_gradient <- c("#FFFFFF", "#81ABD9", "grey80", "#211f82")

ggplot(county_df, aes(long, lat, group = group, fill = pct_dem)) +
  geom_polygon(color = "white", size = 0.1) +
  scale_fill_gradientn(
    colors = my_gradient,
    na.value = "#FFFFFF",
    name = "Percent Democrat"
  ) +
  labs(
    title = "U.S. Presidential Election: Percent Democratic Vote",
    subtitle = "County-level results"
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold")
  )

```

```{r}
library(tidyverse)
library(socviz)

# Load data
election_panel <- read_csv(
  "https://bcdanl.github.io/data/election_panel_2000_2024.csv"
)

# Build dataset for all years
df <- election_panel %>%
  filter(year %in% c(2000, 2004, 2008, 2012, 2016, 2020, 2024))

# Join to county map
county_df <- county_map %>%
  left_join(df, by = c("id" = "fips"))

# Your required 4-color continuous gradient
my_gradient <- c("#FFFFFF", "#81ABD9", "grey80", "#211f82")

# Compute quartiles for legend ticks
q <- quantile(county_df$pct_dem, probs = c(.25, .5, .75), na.rm = TRUE)

ggplot(county_df, aes(long, lat, group = group, fill = pct_dem)) +
  geom_polygon(color = "white", size = 0.05) +
  scale_fill_gradientn(
    colors = my_gradient,
    na.value = "#FFFFFF",
    name = "Percent Democrat",
    limits = range(county_df$pct_dem, na.rm = TRUE),
    breaks = c(min(county_df$pct_dem, na.rm = TRUE), q, max(county_df$pct_dem, na.rm = TRUE)),
    labels = scales::number_format(accuracy = 0.1)
  ) +
  facet_wrap(~ year, ncol = 4) +
  labs(
    title = "U.S. Presidential Elections, 2000–2024",
    subtitle = "Votes toward Democratic Presidential Candidates"
  ) +
  theme_void() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 13),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm")
  )


```

library(tidyverse) library(socviz)

# Load data

election_panel \<- read_csv( "[https://bcdanl.github.io/data/election_panel_2000_2024.csv"](https://bcdanl.github.io/data/election_panel_2000_2024.csv%22) )

# Build dataset for all years

df \<- election_panel %\>% filter(year %in% c(2000, 2004, 2008, 2012, 2016, 2020, 2024))

# Join to county map

county_df \<- county_map %\>% left_join(df, by = c("id" = "fips"))

# Your required 4-color continuous gradient

my_gradient \<- c("#FFFFFF", "#81ABD9", "grey80", "#211f82")

# Compute quartiles for legend ticks

q \<- quantile(county_df\$pct_dem, probs = c(.25, .5, .75), na.rm = TRUE)

ggplot(county_df, aes(long, lat, group = group, fill = pct_dem)) + geom_polygon(color = "white", size = 0.05) + scale_fill_gradientn( colors = my_gradient, na.value = "#FFFFFF", name = "Percent Democrat", limits = range(county_df$pct_dem, na.rm = TRUE),
    breaks = c(min(county_df$pct_dem, na.rm = TRUE), q, max(county_df\$pct_dem, na.rm = TRUE)), labels = scales::number_format(accuracy = 0.1) ) + facet_wrap(\~ year, ncol = 4) + labs( title = "U.S. Presidential Elections, 2000–2024", subtitle = "Votes toward Democratic Presidential Candidates" ) + theme_void() + theme( strip.text = element_text(size = 12, face = "bold"), plot.title = element_text(size = 18, face = "bold"), plot.subtitle = element_text(size = 13), legend.position = "bottom", legend.key.width = unit(2, "cm") )

```{r}
library(tidyverse)
library(socviz)

# Load data
election_panel <- read_csv(
  "https://bcdanl.github.io/data/election_panel_2000_2024.csv"
)

county_map <- socviz::county_map



# Compute summary stats for legend
break_vals <- quantile(
  map_data$pct_DEMOCRAT,
  probs = c(0, 0.25, 0.5, 0.75, 1),
  na.rm = TRUE
)

# Labels like the example
break_labels <- c(
  paste0(round(break_vals[1], 1), "\n(Min)"),
  paste0(round(break_vals[2], 1), "\n(25th)"),
  paste0(round(break_vals[3], 1), "\n(Median)"),
  paste0(round(break_vals[4], 1), "\n(75th)"),
  paste0(round(break_vals[5], 1), "\n(Max)")
)

# Plot
ggplot(map_data, aes(long, lat, group = group, fill = pct_DEMOCRAT)) +
  geom_polygon(color = NA) +
  
  facet_wrap(~year, ncol = 3) +
  
  scale_fill_gradientn(
    colors = c("#FFFFFF", "#211f82", "grey80", "#81ABD9"),
    breaks = break_vals,
    labels = break_labels,
    name = "Percent\nDemocrat",
    na.value = "grey80"
  ) +
  
  coord_quickmap() +
  
  labs(
    title = "U.S. Presidential Elections, 2000–2024",
    subtitle = "Votes toward Democratic Presidential Candidates"
  ) +
  
  theme_void() +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm")
  )
```
